Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2014-07-15 cordillera-cycle.tex
Implement Arjen's comment on intro to results.

% My commands
\newcommand{\note}[1]{\textbf{[NOTE: #1]}}
\newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
\newcommand{\aref}[0]{\textbf{[ref.]}}

@@

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera have been more extensive than today. At the Last Glacial
Maximum (LGM), a continuous blanket of ice, the Cordilleran ice sheet
\citep{Dawson.1888}, extended from the Alaska Range in the north to the
North Cascades in the south. In addition, it extended offshore, where it calved
into the Pacific Ocean, and merged with the western margin of its much larger
neighbour, the Laurentide ice sheet, east of the Rocky Mountains
(Fig.~\ref{fig:locmap}).

@@

\begin{itemize}
  \item How much ice was locked in the Cordilleran ice sheet during the
    last glacial maximum?
  \item Which were its primary dispersal centres? Do they reflect stable or
    ephemeral configurations?
  \item What was the scale of glaciation prior to the last glacial maximum?
  \item How rapid was the last deglaciation? Did it include late glacial
    stand-stills or re-advances?
\end{itemize}

@@

The simulations presented here were run using Parallel Ice Sheet Model (PISM,
development version~11b0a7f and stable version~0.6.1), an open-source,
finite-difference, shallow ice sheet model \citep{PISM-authors.2014}.
The model inputs basal topography, sea level, geothermal
heat flux and climate forcing, and computes the evolution of ice extent
and thickness in time, the thermal and dynamic state of the ice sheet, and
the associated lithospheric response. Our modelling domain encompasses the
entire area covered by the Cordilleran ice sheet at the last glacial maximum
(Fig.~\ref{fig:locmap}).

To reconstruct the successive phases of growth and decay of the last Cordilleran
ice sheet, palaeo-climatic conditions of the last glacial cycle are mimicked
by applying time-dependent temperature offsets derived from multiple
palaeo-temperature proxy records. Each simulation starts from assumed ice-free
conditions at 120\,kyr, and runs to present. These computations were
performed on 16 to 128 cores at the Swedish National Supercomputing Centre.

\subsection{Ice thermodynamics}

PISM is a~shallow model, which implies that the balance of stresses is
approximated based on their predominant components.
The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
Shallow Ice Approximation (SIA) by adding both velocity solutions
\citep[Eqn.~15]{Winkelmann.etal.2011}.
Ice softness depends on temperature and water content through an enthalpy
formulation \citep{Aschwanden.etal.2012}. Surface air
temperature from the atmospheric forcing provides the upper boundary condition
to the ice enthalpy model. Temperature is computed subglacially to a~depth of
3\,km, where it is conditioned by a~geothermal heat flux of
70\,\unit{mW\,m^{-2}}. Although this uniform value does not
account for the high spatial geothermal variability in the region, it is on
average representative of available heat flow measurements
\citep{Blackwell.Richards.2004}.

A~pseudo-plastic sliding law \citep{PISM-authors.2014}\footnote{I need to ask
    Ed Bueler if the sliding will be published somewhere soon.} relates the
bed-parallel shear stress and the sliding velocity. The yield stress is
modelled using the Mohr--Coulomb criterion. The till friction angle $\phi$
varies from 15 to 45{\degree}. It is taken as a~function of modern bed
elevation, with lowest values occurring at low elevation, thus accounting
for a~weakening of till associated with the presence of marine sediments.
The till effective pressure is related to the modelled amount of water in the
till, using a formula derived from laboratory experiments with till extracted
from an Antarctic ice stream \citep[Eqn.~2]{Tulaczyk.etal.2000}. Basal
topography is
derived from the ETOPO1 combined topography and bathymetry dataset with
a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}. Sea level is lowered as
a function of time based on the Spectral Mapping Project (SPECMAP)
\citep[SPECMAP,][]{Imbrie.etal.1989} time scale.
Basal topography responds to ice load
following a bedrock deformation model that includes point-wise isostasy,
elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite
half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.

\subsection{Surface mass-balance}

Ice surface accumulation and ablation are computed from monthly mean
near-surface air temperature, monthly precipitation, and monthly standard
deviation of near-surface temperature by a~temperature-index model
\citep[e.g.,][]{Hock.2003}. Ice accumulation is equal to precipitation
when temperature is below 0\,\unit{{\degree}C}, and decreases to zero linearly
with temperature between 0 and 2\,\unit{{\degree}C}. Ice ablation is computed
from the number of positive degree-days, defined as the integral of
temperatures above 0\,\unit{{\degree}C} in one year.

The positive degree-day integral \citep[Eqn.~6]{Calov.Greve.2005} is
numerically
approximated using week-long sub-intervals. It accounts for temperature
variability assuming a~normal distribution along a~central (input) value. The
temperature standard deviation is part of the forcing climatology. It was
computed from daily temperature values from the North American Regional
Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, in order to account for the
effects of spatial and seasonal variations of temperature variability
\citep[Fig.~\ref{fig:atm};][]{Seguinot.2013}, after excluding variability
associated with the seasonal cycle itself \citep[cf.][]
{Seguinot.Rogozhina.2014}. The ablation model incorporates degree-day factors
of 3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia \citep{Shea.etal.2009}.

\subsection{Atmospheric forcing}

Atmospheric forcing of the model consists of a present-day monthly climatology
$\{T_{m0}, P_{m0}\}$, modified by a lapse-rate correction ${\Delta}T_{LR}$ and
temperature-offset time series ${\Delta}T_{TS}$.
\begin{align}
    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{LR}(t) + {\Delta}T_{TS}(t, x, y) \\
    P_m(t, x, y) &= P_{m0}(x, y)
\end{align}

The present-day climatology $\{T_{m0}, P_{m0}\}$ was computed from
near-surface air temperature and precipitation rate fields of the NARR
\citep{Mesinger.etal.2006} over the period 1979--2000. Modern climate of the
North American Cordillera is characterised by strong geographic variations of
seasonality, timing of the precipitation peak and daily temperature variability
(Fig.~\ref{fig:atm}).
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as identified in a previous sensitivity
study \citep{Seguinot.etal.2014}.

The lapse-rate correction ${\Delta}T_{LR}$ is computed using the NARR surface
geopotential height invariant field as a reference topography $z_{ref}$,
\begin{align}
    {\Delta}T_{LR}(t, x, y) &= -\gamma [z_{s}(t, x, y)-z_{ref}] \\
                            &= -\gamma [h(t, x, y)+z_{b}(t, x, y)-z_{ref}],
\end{align}

thus accounting for the evolution of ice thickness $h$ on the one hand, and
for differences between the the ice flow model basal topography $z_{b}$ and the
NARR topography $z_{ref}$ on the other hand. All simulations use an annual temperature lapse rate of $\gamma = 6\,\unit{{\degree}C\,km^{-1}}$.

Temperature offset time-series ${\Delta}T_{TS}$ are derived from proxy records from
the Greenland Ice Core Project \citep[GRIP,][]{Dansgaard.etal.1993}, the
North Greenland Ice Core Project \citep[NGRIP,][]{Andersen.etal.2004},
the European Project for Ice Coring in Antarctica \citep[EPICA,][]
{Jouzel.etal.2007}, the Lake Vostok ice core \citep{Petit.etal.1999}, and Ocean
Drilling Program (ODP) sites 1012 and 1020, both located off the Californian
shore \citep{Herbert.etal.2001}. Palaeo-temperatures from the GRIP and NGRIP
records were calculated using a quadratic equation \citep{Johnsen.etal.1995},
\begin{equation}
    {\Delta}T_{TS}(t) = -11.88[\chem{\delta^{18}O(t)}-\chem{\delta^{18}O}(0)]
                        -0.1925[\chem{\delta^{18}O(t)}^2-\chem{\delta^{18}O}(0)^2],
\end{equation}
while the temperature reconstructions from the Antarctic and oceanic cores were
provided as such. All records were scaled linearly in
order to simulate realistic and comparable ice extents at the last
glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).

For all proxy records, we firstly run a low-resolution simulation, using
a~10\,km, 300 by 150~points horizontal grid for SSA and SIA velocities,
up to 51~enthalpy layers in the ice, and 31~temperature layers in the bedrock.
For the GRIP and EPICA records, we then run a second simulation using a higher
horizontal resolution of 6\,km, up to 101~ice layers, and 61~bedrock
layers. In the rest of this paper, we occasionally refer to different model
through by the proxy record used for palaeo-temperature forcing.

@@

Despite large differences in input temperature offset time-series, the model
output presents consistent features that can be observed across the range of
forcing data used. In all simulations, total ice volume is relatively low
during most of the glacial cycle, while two major glaciation events occur
during Marine Isotope Stage (MIS)~4 (61.9--55.4\,kyr) and MIS~2
(29.5--16.9\,kyr, Fig.~\ref{fig:timeseries}). A local ice minimum is
consistently attained during MIS~3 (52.2--45.6\,kyr). However, the
magnitude and timing of these three events depend significantly on the choice
of proxy record used for palaeo-temperature forcing.

Simulations forced by the Greenland ice core records (GRIP, NGRIP) produce
highest variability in modelled ice volume throughout the glaciation history.
Model runs using oceanic (ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok)
records generally result in lower
modelled ice volume during MIS~4 and larger ice volume during MIS~3, thus
suggesting lower ice volume variability throughout the simulation length. The
NGRIP palaeo-temperature forcing is the only one that results in larger
glaciation during MIS~4 than during the last glacial maximum (MIS~2).

While simulations driven by the Antarctic and GRIP records
reach a last glacial maximum in ice volume between 19.5 and 16.9\,kyr,
those steered by oceanic and the NGRIP records attain it
several thousands of years earlier. Moreover, the ODP~1012 run yields rapid
deglaciation prior to 20\,kyr due to a local temperature maximum that is
inconsistent with all other records used. The ODP~1020 simulation predicts an
early last glacial maximum in ice volume at 29.5\,kyr, followed by
slower deglaciation than modelled using other records. Finally, model runs
informed by Antarctic ice core records produce a continuous, rapid deglaciation
after the last glacial maximum, whereas the simulations driven by the GRIP
record produce a rapid deglaciation in three steps, including stand-stills and
readvances of the ice margin.

Despite the different timing, snapshots of model output show relatively
consistent patterns of glaciation (Fig.~\ref{fig:snapshots}). As a result of
different scaling factors applied to palaeo-temperature proxy records,
modelled ice sheet geometries at the last glacial maximum appear very similar
from one simulation to the next. They include a central divide located at about
3\,500\,m elevation along the spine of the Rocky Mountains. In all
simulations, a central ice cap persists over the Skeena mountains during MIS~3
between the two glaciation events. However, the size of this ice cap, as well
as the magnitude of the MIS~4 glaciation, depend sensitively on the choice of
palaeo-temperature proxy record used to drive the model.

@@

@@

% tab:records
\begin{table*}[t]
  \caption{Palaeo-temperature proxy records and scaling parameters used to
           prepare temperature offset time-series used to force the ice sheet
           model through the last 120\,kyr. $f$ corresponds to the scaling
           factor used and $T_{[32;22]}$ refers to the mean temperature anomaly
           during the period -32 to~-22~kyr after scaling.}
  \label{tab:records}
  {\begin{tabular}{lccccccl}
    \tophline
    Record & Lat. & Lon. & Elev. & Proxy & $f$ & $T_{[32;22]}$
           & Original reference\\
    \middlehline
    GRIP     &  72{\degree} 35' N  % 72.58 (decimal)
             &  37{\degree} 38' W  % 37.64 (decimal)
             & 3238\,m
             & \chem{\delta^{18}O}
             & 0.35 & -5.8{\degree}C  % -16.4126 (before scaling)
             & \citet{Dansgaard.etal.1993} \\

@@

% fig:locmap
\begin{figure}
  \includegraphics{locmap}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,kyr\,BP} (21.4\,cal\,kyr\,BP)
           \citep{Dyke.2004}. The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           St.-Elias mountains (WSE), the Selwyn and MacKenzie mountains (SMK),
           the Skeena Mountains, the Coast Mountains (CM), the Rocky Mountains
           (RM) and the North Cascades (NC). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.
           \todo{Mark palaeo-ice sheets and mountain ranges on the map.}}
  \label{fig:locmap}
\end{figure}

% fig:atm
\begin{figure}
  \includegraphics{atm}
  \caption{Monthly mean near-surface air temperature, precipitation and
           standard deviation of daily mean temperature for January and July
           months from the North American Regional Reanalysis (NARR)
           climatology, used to force the ice sheet model. Note the
           strong contrasts in seasonality, timing of the precipitation peak,
           and temperature variability over the model domain, notably between
           the maritime and continental regions.}
  \label{fig:atm}
\end{figure}

@@

% fig:snapshots
\begin{figure*}
  \includegraphics{snapshots}
  \caption{Snapshots of modelled surface topography (500\,m contours)
           corresponding to the ice volume extrema indicated on
           Fig.~\ref{fig:timeseries}. Note the occurence of spatial similarities
           despite large differences in timing.}
  \label{fig:snapshots}
\end{figure*}

>>>>

% My commands
\newcommand{\renote}[1]{\footnote{\textbf{Comment reply}: #1}}
\newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
\newcommand{\aref}[0]{\textbf{[ref.]}}

@@

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera have been more extensive than today. At the Last Glacial
Maximum (LGM), a continuous blanket of ice, the Cordilleran ice sheet\renote{
    In \citep{Seguinot.etal.2014}, The Cryosphere changed all my occurences of
    Cordilleran (Greenland etc) \textbf{I}ce \textbf{S}heet to lowercase
    \textbf{i}ce \textbf{s}heet. Let's keep it so for now for consistency
    within the thesis (kappa + papers).}
\citep{Dawson.1888}, extended from the Alaska Range in the north to the
North Cascades in the south. In addition, it extended offshore, where it calved
into the Pacific Ocean, and merged with the western margin of its much larger
neighbour, the Laurentide ice sheet, east of the Rocky Mountains
(Fig.~\ref{fig:locmap}).

@@

\begin{itemize}
  \item How much ice was locked in the Cordilleran ice sheet during the
    last glacial maximum?
  \item What was the scale of glaciation prior to the last glacial maximum?
  \item Which were its primary dispersal centres? Do they reflect stable or
    ephemeral configurations?
  \item How rapid was the last deglaciation? Did it include late glacial
    stand-stills or re-advances?
\end{itemize}

@@

%\subsection{Forcings}

\todo{Move parameter values into a separate table.}

\todo{Do we need a new subsection title here?}

The simulations presented here were run using the Parallel Ice Sheet Model
(PISM, development version~11b0a7f and stable version~0.6.1), an open source,
finite difference, shallow ice sheet model \citep{PISM-authors.2014}. The model
requires input on basal topography, sea level, geothermal heat flux and
atmospheric\renote{
    Changed here from ``climate'' to ``atmospheric''. Although I have used
    ``climate forcing'' in the previous paper, I now think that ``atmospheric
    forcing'' is more appropriate: ``climate'' may also encompass the oceanic
    component, not really assessed here.}
forcing, and computes the evolution of ice extent
and thickness over time, the thermal and dynamic\renote{
    Thermal = temperature, dynamic = force + movement. If I meant movement
    alone, I would use ``cinematic''. Here we could use ``ice temperatures,
    stress and deformation (i.e. velocities)'', but that becomes a bit clumsy,
    I feel.}
states of the ice sheet, and the associated lithospheric response.

Basal topography is derived from the ETOPO1 combined topography and bathymetry
dataset with a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}. Sea level
is lowered as a function of time based on the Spectral Mapping Project\renote{
    I had no time to check for later products, but if you have suggestions for
    something more up-to-date, I can try to include it in later simulations
    (everything will have to be re-run anyway).}
\citep[SPECMAP;][]{Imbrie.etal.1989} time scale. Geothermal heat flux
is applied as a constant value of 70\,\unit{mW\,m^{-2}} at 3\,km depth
(Sect.~\ref{sec:icedyn}). Atmospheric forcing is provided by a monthly
climatology from the North American Regional Reanalysis
\citep[NARR;][]{Mesinger.etal.2006} perturbated by time-dependent and
lapse-rate temperature corrections (Sect.~\ref{sec:atm}).

Each simulation starts from assumed ice-free conditions at 120\,000 years ago
(120\,ka), and runs to present. Our modelling domain of 1\,500 by 3\,000\,km
encompasses the entire area covered by the Cordilleran ice sheet at the LGM
(Fig.~\ref{fig:locmap}). The simulations were run on two distinct grids, using
a lower horizontal resolution of 10\,km, and a higher horizontal resolution of
6\,km. These computations were performed on 16 to 128 computing cores at the
Swedish National Supercomputing Centre.

\subsection{Ice thermodynamics}
\label{sec:icedyn}

Ice sheet dynamics are typically modelled using a combination of internal
deformation and basal sliding. PISM is a~shallow ice sheet model, which implies
that the balance of stresses is approximated based on their predominant
components. The Shallow Shelf Approximation (SSA) is used as a ``sliding law''
for the Shallow Ice Approximation (SIA) by adding both velocity solutions
\citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}\renote{
    Arjen, I don't agree with your comment. Reference to equation number(s)
    show the reader where to look. \citet{Winkelmann.etal.2011} is a rather
    long paper that also treats many other aspects of the model. Their SIA+SSA
    approach builds on \citep{Bueler.Brown.2009}, but it is slightly different.
    It is the one in use in PISM, and in my simulations.}.
Ice rheology depends on temperature and water content through an enthalpy
formulation \citep{Aschwanden.etal.2012}. Surface air temperature derived from
the atmospheric forcing (Sect.~\ref{sec:atm}) provides the upper boundary
condition to the ice enthalpy model. Temperature is computed subglacially to
a~depth of 3\,km\renote{
    Yes, I changed this from 1 to 3\,km as we now model longer period of times
    than in the previous paper. The penetration depth of temperature anomalies
    (i.e. skin depth) depends on the period of temperature oscillations at the
    surface. The longer the period (in our case 100\,kyr), the deeper the
    temperature change. A good way to estimate what this parameter should be is
    to solve the 1-D temperature diffusion equation in the case of periodic
    surface forcing, using rock heat capacity and rock thermal conductivity
    values similar as in the numerical model. This basically forms the basis
    for my choice of the 3\,km value.},
where it is conditioned by a lower boundary geothermal heat flux of
70\,\unit{mW\,m^{-2}}. Although this uniform value does not account for the
high spatial geothermal variability in the region
\citep{Blackwell.Richards.2004}, it is, on average, representative of available
heat flow measurements. In the low-resolution simulations, the vertical grid
consists of to 51~enthalpy layers in the ice sheet and 31~temperature layers in
the bedrock. In the high-resolution simulations, up to 101~ice layers, and
61~bedrock layers are used.

A~pseudo-plastic sliding law \citep{Bueler.Pelt.2014} relates the
bed-parallel shear stress to the sliding velocity. The yield stress, $\tau_c$,
is modelled using the Mohr--Coulomb criterion,
\begin{equation}
   \tau_c = c_0 + N\,\tan{\phi} \,,
\end{equation}
where cohesion, $c_0$, is assumed to be zero. The friction angle, $\phi$,
varies from 15 to 45{\degree}\renote{
    Different from the previous paper, but fairly similar in effect (see
    $\alpha$ and $\delta$ parameter values and corresponding equations in the
    kappa, basal sliding section). These values are is based on a quick
    sensitivity test on Greenland.}.
It is taken to be a~function of modern bed elevation, with lower values at
lower elevations (<200\,m above sea level), thus accounting for a~weakening of
till associated with the presence of marine sediments. Effective pressure, $N$,
is related to the ice overburden stress, $\rho gh$, and the modelled amount of
subglacial water, using a formula derived from laboratory experiments with till
extracted from an Antarctic ice stream \citep{Tulaczyk.etal.2000,
Bueler.Pelt.2014},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
\end{equation}
where $delta$ is choosed as 0.95, $e_0$ is a measured reference void ratio and
$C_c$ a measured compressibility coefficient. The amount of water at the base,
$W$, varies from zero to $W_{max}=2$\,m, a threshold above which instantaneous
drainage is assumed\renote{
    Regarding a parameter table, it is in the kappa, I am skipping it for now
    but will try to include it soon.}.
Finally, the bedrock topography responds to ice load\renote{
    I disagree with writing ``the isostatic response''. The main point in
    using this model is that computes a non-isostatic response.}
following a bedrock deformation model that includes point-wise isostasy,
elastic lithosphere flexure and viscous mantle deformation in the lower
half-space\renote{
    I had put an emphasis on semi-infinite at this is the key assumption of
    that model, allowing effective solution in Fourier domain. This is the
    whole point of Ed's paper and probably the reason why he picked it for PISM
    But yes, I guess that ``lower half-space'' is fine, too...}
\citep{Lingle.Clark.1985,Bueler.etal.2007}.

\subsection{Surface mass balance}

\renote{
    Surface mass balance is the balance of mass fluxes at the surface,
    consisting of ice accumulation and ice ablation, which I introduce straight
    away. I don't know what kind of ``opening sentence'' can be added here.}
Ice surface accumulation and ablation are computed from monthly mean
near-surface air temperature $T_m$, monthly standard deviation of near-surface
air temperature $\sigma$, and monthly precipitation $P_m$, using
a~temperature-index model \citep[e.g.,][]{Hock.2003}. Accumulation is equal to
precipitation when air temperatures are below 0\,\unit{{\degree}C}, and
decreases to zero linearly with temperatures between 0 and
2\,\unit{{\degree}C}. Ablation is computed from the number of positive
degree-days (PDD), defined as the integral of temperatures above
0\,\unit{{\degree}C} in one year.

The PDD computation accounts for stochastic temperature variations by assuming
a normal temperature distribution of standard deviation $\sigma$ aroung the
expected value $T_m$. It is expressed by an error-function formulation
\citep{Calov.Greve.2005},
\begin{equation}
    \label{eqn:calovgreve}
    \mathrm{PDD} = \int_{t_1}^{t_2} \mathrm{d}t
        \left[\frac{\sigma}{\sqrt{2\pi}}
                \exp\left({-\frac{T_{m}^2}{2\sigma^2}}\right)
              + \frac{T_{m}}{2} \, \mathrm{erfc}
                \left(-\frac{T_{m}}{\sqrt{2}\sigma}\right)\right] \,,
\end{equation}
which is numerically approximated using week-long sub-intervals. In order to
account for the effects of spatial and seasonal variations of temperature
variability \citep{Seguinot.2013}, $\sigma$ is computed from daily temperature
values from the North American Regional Reanalysis
\citep[NARR,][]{Mesinger.etal.2006}, after excluding variability associated
with the seasonal cycle itself \citep[cf.][]{Seguinot.Rogozhina.2014}\renote{
    I may add a figure in the kappa if I find time, but I don't think that we
    should do it here. The subject is too marginal for the present study.}.
The ablation model incorporates degree-day factors of
3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia \citep{Shea.etal.2009}.

\subsection{Atmospheric forcing}
\label{sec:atm}

Atmospheric forcing of the model consists of a present-day monthly climatology,
$\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
${\Delta}T_{TS}$, and lapse-rate corrections, ${\Delta}T_{LR}$:
\begin{align}
    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t)
                    + {\Delta}T_{LR}(t, x, y) \,, \\
    P_m(t, x, y) &= P_{m0}(x, y) \,.
\end{align}

The present-day monthly climatology, $\{T_{m0}, P_{m0}\}$, was computed from
near-surface air temperature and precipitation rate fields of the NARR
\citep{Mesinger.etal.2006}, averaged between 1979 and 2000. Modern climate of the
North American Cordillera is characterised by strong geographic variations in
temperature seasonality, timing of the maximum annual precipitation, and
daily temperature variability (Fig.~\ref{fig:atm}).
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as identified in a previous sensitivity
study \citep{Seguinot.etal.2014}.

Temperature offset time-series, ${\Delta}T_{TS}$, are derived from
palaeo-temperature proxy records from
the Greenland Ice Core Project \citep[GRIP,][]{Dansgaard.etal.1993}, the
North Greenland Ice Core Project \citep[NGRIP,][]{Andersen.etal.2004},
the European Project for Ice Coring in Antarctica \citep[EPICA,][]
{Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and Ocean
Drilling Program (ODP) sites 1012 and 1020, both located off the Californian
coast \citep{Herbert.etal.2001}. Palaeo-temperatures from the GRIP and NGRIP
records were calculated from oxygen isotope (\chem{\delta^{18}O}) measurements
using a quadratic equation \citep{Johnsen.etal.1995},
\begin{equation}
    {\Delta}T_{TS}(t) = -11.88 [\chem{\delta^{18}O}(t)
                               -\chem{\delta^{18}O}(0)]
                        -0.1925[\chem{\delta^{18}O}(t)^2
                                -\chem{\delta^{18}O}(0)^2] \,,
\end{equation}
while temperature reconstructions from Antarctic and oceanic cores were
provided as such. All records were scaled linearly in
order to simulate realistic and comparable ice limits at the last
glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).
In the rest of this paper, we occasionally refer to different model
runs by the name of the proxy record used for the palaeo-temperature forcing.

Finally, lapse-rate corrections, ${\Delta}T_{LR}$, are computed as a function
of ice surface elevation, $s$, using the NARR surface geopotential height
invariant field as a reference topography, $b_{ref}$, by
\begin{align}
    {\Delta}T_{LR}(t, x, y) &= -\gamma [s(t, x, y)-b_{ref}] \\
                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{ref}],
\end{align}

thus accounting for the evolution of ice thickness, ${h=s-b}$\renote{
    Small change in notation to match with the kappa},
on the one hand, and
for differences between the the ice flow model basal topography, $b$, and the
NARR reference topography, $b_{ref}$, on the other hand. All simulations use an
annual temperature lapse rate of $\gamma = 6\,\unit{{\degree}C\,km^{-1}}$.

@@

\todo{Add table with MIS~2, 3, and 4 volume and area table.}

Despite large differences in the input atmospheric forcing (temperature offset
time-series; Fig.~\ref{fig:timeseries}), model output presents consistent
features that can be observed across the range of forcing data used. In all
simulations, modelled ice volumes remain relatively low during most of the
glacial cycle, except during two major glacial events which occur between 61.9
and 55.4\,ka during Marine Oxygen Isotope Stage (MIS)~4, and between 29.5 and
16.9\,ka during MIS~2 (Fig.~\ref{fig:timeseries}). An ice volume minimum is
consistently attained between 52.2 and 45.6\,ka during MIS~3. However, the
magnitude and precise timing of these three events depend significantly on the
choice of proxy record used for time-dependent atmospheric forcing.

Simulations forced by the Greenland ice core palaeo-temperature\renote{
    Throughout this section, I prepended ``palaeo-temperature'' to
    ``record(s)'' in place of ``climate'' as you originally suggested. If we
    are to precise what type ``records'' we use, let's be clear about it.}
records (GRIP, NGRIP) produce the highest variability in modelled ice volume
throughout the last glacial cycle. In contrast, simulations driven by oceanic
(ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok) palaeo-temperature records
generally result in smaller\renote{
    Until we run out of other ajectives, I don't think that ``modest'' and
    ``generous'' are adequate terms for describing ice volumes.}
modelled ice volumes during MIS~4 and larger ice volumes during MIS~3, thus
indicating. More broadly speaking, they produce lower ice volume variability
throughout the simulation length. The NGRIP atmospheric forcing is the only one
that results in a larger ice volume during MIS~4 than during the last glacial
maximum (MIS~2).

While simulations driven by GRIP and the two Antarctic palaeo-temperature
records attain a last ice volume maximum between 19.5 and 16.9\,ka, those
informed by NGRIP and the two oceanic palaeo-temperature records attain their
maximum ice volume thousands of years earlier. Moreover, the ODP~1012 run
yields a rapid deglaciation of the modelled area prior to 20\,ka. The ODP~1020
simulation predicts an early maximum in ice volume at 29.5\,ka, followed by
slower deglaciation than modelled using other palaeo-temperature records.
Finally, model runs forced by Antarctic palaeo-temperature records produce a
rapid an uninterrupted deglaciation after the last glacial maximum, whereas the
simulation driven by the GRIP palaeo-temperature record also yields a rapid
deglaciation but in three steps, interrupted by two stand-stills and readvances
of the ice margin.

Despite such differences in the timing of attained volume maxima and mimima,
snapshots of model output show relatively consistent patterns of glaciation
(Fig.~\ref{fig:snapshots}). As a result of scaling factors applied to the
different palaeo-temperature proxy records (Table~\ref{tab:records}), modelled
ice sheet geometries during the last glacial maximum (MIS~2;
Fig.~\ref{fig:snapshots}, lower panels) appear highly similar from one
simulation to the next. They include a ca. 1\,500\,km-long central divide
located about 3\,500\,m\ above sea level (a.s.l.) along the spine of the Rocky
Mountains. A central ice cap persists in all simulations over the Skeena
Mountains during the MIS~3 ice volume minima (Fig.~\ref{fig:snapshots}, middle
panels). However, the size of the Skeena Mountains ice cap, and the dimensions
of the Cordilleran ice sheet during the MIS~4 glaciation, depend sensitively on
the choice of palaeo-temperature record used to drive the model.

@@

\todo{Add some point, refer back to Fig.~\ref{fig:atm}.}

@@

% tab:records
\begin{table*}[t]
  \caption{Palaeo-temperature proxy records and scaling parameters yielding
           temperature offset time-series used to force the ice sheet model
           through the last glacial cycle (Fig.~\ref{fig:timeseries}). $f$
           corresponds to the scaling factor adopted to yield last glacial
           maximum ice limits in the vicinity of mapped end moraines, and
           $T_{[32, 22]}$ refers to the resulting mean temperature anomaly
           during the period -32 to~-22~ka after scaling.}
  \label{tab:records}
  {\begin{tabular}{lccccccl}
    \tophline
    Record & Latitude & Longitude & Elev. & Proxy & $f$ & $T_{[32, 22]}$
           & Reference\\
    & & & (m\,a.s.l) & & & (K) & \\
    \middlehline
    GRIP     &  72{\degree} 35' N  % 72.58 (decimal)
             &  37{\degree} 38' W  % 37.64 (decimal)
             & 3238\,m
             & \chem{\delta^{18}O}
             & 0.35 & -5.8{\degree}C  % -16.4126 (before scaling)
             & \citet{Dansgaard.etal.1993} \\

@@

% fig:locmap
\begin{figure}
  \includegraphics{locmap}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,kyr\,BP} (21.4\,cal\,kyr\,BP)
           \citep{Dyke.2004}. The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           St.-Elias mountains (WSE), the Selwyn and MacKenzie mountains (SMK),
           the Skeena Mountains (SM), the Coast Mountains (CM), the Rocky
           Mountains (RM) and the North Cascades (NC). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.
           \todo{Mark palaeo-ice sheets and mountain ranges on the map.}}
  \label{fig:locmap}
\end{figure}

% fig:atm
\begin{figure}
  \includegraphics{atm}
  \caption{Monthly mean near-surface air temperature, precipitation and
           standard deviation of daily mean temperature for January and July
           months from the North American Regional Reanalysis (NARR)
           climatology, used to force the ice sheet model. Note the
           strong contrasts in seasonality, timing of the precipitation peak,
           and temperature variability over the model domain, notably between
           the maritime and continental regions.
           \todo{Dash boundaries between ice sheets?}}
  \label{fig:atm}
\end{figure}

@@

% fig:snapshots
\begin{figure*}
  \includegraphics{snapshots}
  \caption{Snapshots of modelled surface topography (500\,m contours)
           corresponding to the ice volume extrema indicated on
           Fig.~\ref{fig:timeseries}. Note the occurence of spatial similarities
           despite large differences in timing.
           \todo{Perhaps indicate the location of the Skeena Mountains.}}
  \label{fig:snapshots}
\end{figure*}
